A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-07-22 manuscript.tex
feedback Rei

To illustrate the functioning and utility of feature-based time series clustering with psychological ESM data, we apply the clustering process to a recent set of studies that collected data on migration experiences. Researchers have recently started using ESM data to follow the daily interactions of migrants with the cultural majority groups \citep[e.g.,][]{Keil2020}. This type of research comes as a response to prominent reviews within the migration literature, which have called for more longitudinal \citep[e.g.,][]{Ward2019} and real-world data \citep[e.g.,][]{McKeown2017}, while at the same time, conceptual works have pushed for more diverse assessments of migration experiences, including motivational, affective, cognitive, and behavioral aspects \citep[e.g.,][]{Kreienkamp2022d}. Importantly for our illustration, the migration ESM research, thus, exemplifies some of the real-world data issues that ESM data commonly faces, including a multivariate conceptualization with event-specific missingness patterns.

@@

To introduce and illustrate the feature-based clustering analysis, we will follow the conceptual steps of the procedure in sequential order and discuss key issues for each step. To do so we will follow the common separation that has structured feature-based clustering into four main steps \citep{rasanen2009, wang2006}. (1) The selection and preparation of the input variables, (2) the extraction of the features that describe the time series, with an optional feature reduction step if there are too many data points for the clustering algorithms, (3) the actual clustering of the time series features, and (4) the evaluation and interpretation of the clusters. 
While this is a common conceptual separation of procedural elements, it is important to note that these steps are a general outline, and the specific details of the analysis will depend on the nature of the data and the research question being addressed. Nonetheless, the conceptual nature of these steps allows us to introduce the mayor elements of the analysis. We also provide a conceptual overview that can be used alongside this section in \fgrref{fig:TSCFlow}.

@@

Importantly, we show that feature-based clustering is not only approachable but provides interpretable and transparent insights about the grouped patterns. For our example of migration experiences, the method was useful to discern adaptive from more stressful experiences and helped to contextualize divergent experiences. We found that some variables, such as interaction quality perceptions or need fulfillment were particularly important in distinguishing the groups (see \fgrref[A]{fig:clusterFeatVar}). Similarly, we found that the central tendency (median), variability (MAD), and linear trend (slope) were the most impactful dynamic features in discerning the trajectory clusters (see \fgrref[B]{fig:clusterFeatVar}). Jointly these two approaches allowed us to identify a cluster that had generally positive and improving experiences while the other cluster had more negative and deteriorating experiences. We were even able to further contextualize the results with out-of-feature comparisons, where we found that the group with the more difficult experiences also reported substantially more discrimination experiences during the post-test (see, e.g., \fgrref[B]{fig:clusterFeatVar}). In short, the feature-based approach allowed us to identify directly interpretable and meaningful groups, where we transparently know what data input the clusters are based on.

@@

As an illustration, the choice of features to extract from the time series data is a critical step that can significantly influence the results of the clustering process. In the current example, we chose to extract features such as means, autocorrelations, and linear trends, which are psychologically and conceptually meaningful in interpreting our time series clusters. However, this selection is not exhaustive and may not capture all relevant aspects of the time series data. For example, we did not consider attributes like periodicity or spectral density, which could shed light on the data's cyclical patterns. The choice of features largely hinges on the researcher's specific research question and assumptions about the data, thereby injecting a level of subjectivity into the process. Similar challenges arise with the choice of the clustering algorithm or the cluster illustration. These challenges are not unique to feature-based clustering, rather they are common to all clustering approaches \citep{liao2005}. However, it is important to remember that multi-stepped data-driven approaches are particularly vulnerable to the impact of the researchers' degrees of freedom.

@@

For instance, given that the approach does not assume the stationarity restrictions of many model-based approaches, future research can now more easily integrate many of the (non-)linear trend features. Research on capturing nonlinear trends has been growing over the past years and there are exciting possibilities to bring these developments to ESM data \citep{bringmann2023}. For example, bicoherence metrics, polynomial-, and differential equation parameters may be used to capture the type and structure of nonlinear developments \citep[e.g. shape-based approaches in \fgrref{fig:tsClustTax}; see also][]{caro-martin2018, mayor2022}. New features capturing nonlinear structures would add to the under-studied trend features of ESM data.

@@

Beyond the direct academic use, the feature-based time series clustering approach also addresses practical and applied uses. For practitioners, the approach offers a practical and grounded method for dealing with the challenges of complex and messy patient-, customer-, or user data. Not only does the approach directly deal with dimensionality, missingness, and time scales in the time series, but the interpretability and transparency aspects offer particular utility in applied settings, where the costs of misspecification are high. Additionally, the approach is also more readily accessible to practitioners who may not have extensive training in complex data analysis techniques. We provide practical algorithm overviews and readily available code for data preparation, analysis, and interpretation. The ability to identify and interpret meaningful patterns in time series data can have significant implications for practice, particularly in fields such as clinical, organizational, or social psychology, where understanding individual differences and developmental patterns can inform interventions and decision-processes.

>>>>

To illustrate the functioning and utility of feature-based time series clustering with psychological ESM data, we apply the clustering process to a recent set of studies that collected data on migration experiences. Researchers have recently started using ESM data to follow the daily interactions of migrants with the cultural majority groups \citep[e.g.,][]{Keil2020}. This type of research comes as a response to prominent reviews within the migration literature, which have called for more longitudinal \citep[e.g.,][]{Ward2019} and real-world data \citep[e.g.,][]{McKeown2017}, while at the same time, conceptual works have pushed for more diverse assessments of migration experiences, including motivational, affective, cognitive, and behavioral aspects \citep[e.g.,][]{Kreienkamp2022d}. Importantly for our illustration, the migration ESM research, thus, exemplifies some of the real-world data issues that ESM data commonly face, including a multivariate conceptualization with event-specific missingness patterns.

@@

To introduce and illustrate the feature-based clustering analysis, we will follow the conceptual steps of the procedure in sequential order and discuss key issues for each step. To do so we will follow the common separation that has structured feature-based clustering into four main steps \citep{rasanen2009, wang2006}. (1) The selection and preparation of the input variables, (2) the extraction of the features that describe the time series, with an optional feature reduction step if there are too many data points for the clustering algorithms, (3) the actual clustering of the time series features, and (4) the evaluation and interpretation of the clusters. 
While this is a common conceptual separation of procedural elements, it is important to note that these steps are a general outline, and the specific details of the analysis will depend on the nature of the data and the research question being addressed. Nonetheless, the conceptual nature of these steps allows us to introduce the major elements of the analysis. We also provide a conceptual overview that can be used alongside this section in \fgrref{fig:TSCFlow}.

@@

Importantly, we show that feature-based clustering is not only approachable but provides interpretable and transparent insights about the grouped patterns. For our example of migration experiences, the method was useful to discern adaptive from more stressful experiences and helped to contextualize divergent experiences. We found that some variables, such as interaction quality perceptions or need fulfillment were particularly important in distinguishing the groups (see \fgrref[A]{fig:clusterFeatVar}). Similarly, we found that the central tendency (\textit{median}), variability (\textit{MAD}), and linear trend (\textit{slope}) were the most impactful dynamic features in discerning the trajectory clusters (see \fgrref[B]{fig:clusterFeatVar}). Jointly these two approaches allowed us to identify a cluster that had generally positive and improving experiences while the other cluster had more negative and deteriorating experiences. We were even able to further contextualize the results with out-of-feature comparisons, where we found that the group with the more difficult experiences also reported substantially more discrimination experiences during the post-test (see, e.g., \fgrref[B]{fig:clusterFeatVar}). In short, the feature-based approach allowed us to identify directly interpretable and meaningful groups, where we transparently know what data input the clusters are based on.

@@

As an illustration, the choice of features to extract from the time series data is a critical step that can significantly influence the results of the clustering process. In the current example, we chose to extract features such as means, autocorrelations, and linear trends, which are psychologically and conceptually meaningful in interpreting our time series clusters. However, this selection is not exhaustive and may not capture all relevant aspects of the time series data. For example, we did not consider attributes like periodicity or spectral density, which could shed light on the data's cyclical patterns. The choice of features largely hinges on the researcher's specific research question and assumptions about the data, thereby injecting a level of subjectivity into the process. Similar challenges arise with the choice of the clustering algorithm or the cluster illustration. These challenges are not unique to feature-based clustering, rather they are common to all clustering approaches \citep{liao2005, horne2020}. However, it is important to remember that multi-stepped data-driven approaches are particularly vulnerable to the impact of the researchers' degrees of freedom.

@@

For instance, given that the approach does not assume the stationarity restrictions of many model-based approaches, future research can now more easily integrate many of the (non-)linear trend features. Research on capturing nonlinear trends has been growing over the past years and there are exciting possibilities to bring these developments to ESM data \citep{bringmann2023}. For example, bicoherence metrics, polynomial-, and differential equation parameters may be used to capture the type and structure of nonlinear developments \citep[e.g. shape-based approaches in \fgrref{fig:tsClustTax}; see also][]{caro-martin2018, mayor2022}. New features capturing nonlinear structures would add to the under-studied (non-)linear trend features of ESM data.

@@

Beyond the direct academic use, the feature-based time series clustering approach also addresses practical and applied uses. For practitioners, the approach offers a practical and grounded method for dealing with the challenges of complex and messy data from multiple patients, customers, or users. Not only does the approach directly deal with dimensionality, missingness, and time scales in the time series, but the interpretability and transparency aspects offer particular utility in applied settings, where the costs of misspecification are high. Additionally, the approach is also more readily accessible to practitioners who may not have extensive training in complex data analysis techniques. We provide practical algorithm overviews and readily available code for data preparation, analysis, and interpretation. The ability to identify and interpret meaningful patterns in time series data can have significant implications for practice, particularly in fields such as clinical, organizational, or social psychology, where understanding individual differences and developmental patterns can inform interventions and decision-processes.

### 2023-07-22 sections/01DataDescription.Rmd
feedback Rei

The data set we use consists of three studies that followed migrants who had recently arrived in the Netherlands in their daily interactions with the Dutch majority group members \citep[for the data set see][]{Kreienkamp2022b}. After a general migration-focused pre-questionnaire, participants were invited twice per day to report on their (potential) interactions with majority group members for at least 30 days. The short ESM surveys were sent out at around lunch (12pm) and dinner time (7pm). After the 30 day study period participants filled in a post-questionnaire that mirrored the pre-questionnaire. Participants received either monetary compensation or partial course credits based on the number of surveys they completed.

The original studies included `r sum(missS123$pptFull)` participants ($N_{S1}=$ `r missS123$pptFull[missS123$study==1]`, $N_{S2}=$ `r missS123$pptFull[missS123$study==2]`, $N_{S3}=$ `r missS123$pptFull[missS123$study==3]`) with a total of `r format(sum(missS123$nFull), big.mark=",")` ESM measurements. Each of the studies focused on recently arrived first-generation migrants and each study included a number of ideosyncratic variables relevant for the broader research collective. For our empirical example we focus on the variables that were collected during the ESM surveys and were available in all three studies. Variable selection and preparation is described as part of the illustration below but for additional methodological details about the study setup see \citet[][]{Kreienkamp2022b}.

>>>>

The data set we use consists of three studies that followed migrants who had recently arrived in the Netherlands in their daily interactions with the Dutch majority group members \citep[for the data set see][]{Kreienkamp2022b}. After a general migration-focused pre-questionnaire, participants were invited twice per day to report on their (potential) interactions with majority group members for at least 30 days. The short ESM surveys were sent out at around lunch (12pm) and dinner time (7 pm). After the 30-day study period participants filled in a post-questionnaire that mirrored the pre-questionnaire. Participants received either monetary compensation or partial course credits based on the number of surveys they completed.

The original studies included `r sum(missS123$pptFull)` participants ($N_{S1}=$ `r missS123$pptFull[missS123$study==1]`, $N_{S2}=$ `r missS123$pptFull[missS123$study==2]`, $N_{S3}=$ `r missS123$pptFull[missS123$study==3]`) with a total of `r format(sum(missS123$nFull), big.mark=",")` ESM measurements. Each of the studies focused on recently arrived first-generation migrants and each study included a number of idiosyncratic variables relevant for the broader research collective. For our empirical example, we focus on the variables that were collected during the ESM surveys and were available in all three studies. Variable selection and preparation are described as part of the illustration below but for additional methodological details about the study setup see \citet[][]{Kreienkamp2022b}.

### 2023-07-22 sections/02InputVariables.Rmd
feedback Rei

Once the important variables have been selected, the data needs to be prepared for the analysis steps. Importantly, this not only means validating and cleaning the data (e.g., re-coding, removing duplicate or unwanted measurements) but also making the time series comparable. Two important steps are making the time-frames and response scales comparable across participants --- for example, by choosing a time frame that is common to most participants and standardizing the participants' responses \citep{liao2005}.

In our illustration data set, the studies differed substantially in the maximum length of participation ($\text{max}(t_{S1})=$ `r missS123$timeFull[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeFull[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeFull[missS123$study==3]`). This was likely due to the option to continue participation without compensation in the later two studies. To make the three studies comparable in participation and time frames, we iteratively removed all measurement occasions and participants that had more than 45% missingness \citep[which was in line with the general rcecommendation for data that might still need to rely on imputations for later model testing][]{Madley-Dowd2019}. This procedure let to a final sample of `r sum(missS123$pptRed)` participants, who jointly produced `r format(sum(missS123$nRed), big.mark=",")` measurements. Importantly, both the participant repsonse-patterns and the time frame were now substantially more comparable ($\text{max}(t_{S1})=$ `r missS123$timeRed[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeRed[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeRed[missS123$study==3]`). Full methodological details are available in Online Supplemental Material A, but basic item information, descriptives, and correlations are also available in \tblref{tab:descrLong}.

>>>>

Once the important variables have been selected, the data needs to be prepared for the analysis steps. Importantly, this not only means validating and cleaning the data (e.g., re-coding, removing duplicate or unwanted measurements) but also making the time-series comparable. Two important steps are making the time-frames and response scales comparable across participants --- for example, by choosing a time frame that is common to most participants and standardizing the participants' responses \citep['data exclusion' and 'data transformation' in \fgrref{fig:TSCFlow}; also see][]{liao2005}.

In our illustration data set, the studies differed substantially in the maximum length of participation ($\text{max}(t_{S1})=$ `r missS123$timeFull[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeFull[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeFull[missS123$study==3]`). This was likely due to the option to continue participation without compensation in the latter two studies. To make the three studies comparable in participation and time frames, we iteratively removed all measurement occasions and participants that had more than 45% missingness \citep[which was in line with the general recommendation for data that might still need to rely on imputations for later model testing][]{Madley-Dowd2019}. This procedure led to a final sample of `r sum(missS123$pptRed)` participants, who jointly produced `r format(sum(missS123$nRed), big.mark=",")` measurements. Importantly, both the participant response-patterns and the time frame were now substantially more comparable ($\text{max}(t_{S1})=$ `r missS123$timeRed[missS123$study==1]`, $\text{max}(t_{S2})=$ `r missS123$timeRed[missS123$study==2]`, $\text{max}(t_{S3})=$ `r missS123$timeRed[missS123$study==3]`). Full methodological details are available in Online Supplemental Material A, but basic item information, descriptives, and correlations are also available in \tblref{tab:descrLong}.

### 2023-07-22 sections/03FeatureExtraction.Rmd
feedback Rei

\paragraph{Central tendency.}
The central tendency refers to the statistical measures that represent the "typical" or "average" of a set of data. The most common measures of central tendency are the mean, median, and mode \citep{weisberg1992}. As a familiar statistic from probability theory, the central tendency sits at the heart of many fundamental questions about psychological time series. Researchers might, for example, be interested in whether "Over a one month period, are some people happier than others?"

@@

\paragraph{Self-similarity.}
Self-similarity in time series data refers to the property of a time series to exhibit similar patterns of behavior over different time scales \citep{dmello2021}. That is, self-similarity describes how much a measurement carries over to future measurements. One important self-similarity in psychological time series is \texit{inertia} --- how much a measurement carries over to its next measurement \citep{kuppens2010, suls1998}. If inertia is high a development tends to stay in a certain state. Because high inertia is resistant to change, in emotion dynamics high inertia of negative affect has been found to be indicative of underreactive systems and to be characteristic of psychological maladjustment \citep{kuppens2010}. In a similar vein, high inertia in negative affect at baseline was even predictive of the initial onset of depression \citep{kuppens2012}. Conceptually, inertia is more broadly connected to research questions such as: ``Do patients stay in a negative mood for several measurements?''

@@

\paragraph{Linear trend.}
In non-stationary time series a linear trend can be observed when there is a consistent increase or decrease in the data over time \citep{nyblom1986}. For psychological time series, researchers have, for example, pointed out the importance of linear trends in interpersonal communications \citep{vasileiadou2014}, and emotion dynamics \citep{oravecz2016}. Theoretically, linear trends are often considered the simplest way of assessing whether a psychological theory of change is appropriate \citep{gottman1969}. In empirical practice, linear trends are, thus, commonly exemplified by research questions such as "Do patient symptoms improve consistently?" or "Does worker productivity decline continuously?"

For the variables in our illustration data set we chose an overall linear regression slope to capture the linear trend. The regression slope $b_{ij}$ provides the average change from one time point $t$ to the next across all measurements $x$ of a time series $X$ of participant $i$ and variable $j$. The specific form of the OLS slope formula we provide below, calculates $b_{ij}$ as the sum across all time points of the product of the deviation of time $t$ from its mean $\overline{t}$ and the deviation of $x_{ij}$ from its mean $\overline{x}_{ij}$ at each time point, divided by the sum across all time points of the square of the deviation of time from its mean ($\sum(t-\overline{t})^2$). Intuitively, the formula captures the rate of change of variable $x_{ij}$ with respect to time. This slope will indicate how the variable $x_{ij}$ changes over time, controlling for its mean value and the mean of time. If the slope is positive, $x_{ij}$ increases over time; if it's negative, $x_{ij}$ decreases over time.

@@

\paragraph{Nonlinearity.}
Changes in psychology are not always linear, instead nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. Conceptually, researchers might have research question about the type of the development: "Is the development of anxiety a nonlinear process?" as well as the shape and structure of the development: "How many spikes in anxiety did a patient experience?" 
We summarized the nonlinear trend with the \textit{estimated degrees of freedom} of an empty GAM spline model. The $edf$ summarizes the \textit{wiggliness} of a spline trend line by estimating the number of parametric and non-parametric (or smooth) terms $p$ in the model. An edf of 1 would be equivalent to a linear relationship (i.e., one linear slope parameter), whereas a higher edf (particularly an edf > 2) is indicative of a non-linear trend.

@@

Beyond our main features of interest, we also extracted the participant's number of completed ESM measurements to ensure that the clusters are comparable in that regard (i.e., to exclude spurious explanations for the cluster assignments). After the feature extraction, we found that about `r format(round(pMissFull$pmiss %>% filter(set == "All") %>% select(pMissTotal) %>% pull, 2), nsmall=2)`% of the extracted features are missing across the `r ncol(featFull$features %>% select(-ends_with(featureExclusion)))-1` features per participant. This might, for example, happen if participants do not have two subsequent measurements with outgroup interactions, so that an autocorrelation with lag-1 cannot be calculated for the contact-specific variables. The small number of missing values indicates that the feature-based approach indeed largely avoids the structural missingness issue. However, even the few missing values can be an issue for some feature reduction or feature clustering algorithms. We, thus, impute the missing feature values with a single predictive mean matching imputation using the MICE library \citep[][]{buuren2011}.

>>>>

\paragraph{Central tendency.}
The central tendency refers to the statistical measures that represent the "typical" or "average" of a set of data. The most common measures of central tendency are the mean, median, and mode \citep{weisberg1992}. As a familiar statistic from probability theory, the central tendency sits at the heart of many fundamental questions about psychological time series. Researchers might, for example, be interested in whether "Over a one-month period, are some people happier than others?"

@@

\paragraph{Self-similarity.}
Self-similarity in time series data refers to the property of a time series to exhibit similar patterns of behavior over different time scales \citep{dmello2021}. That is, self-similarity describes how much a measurement carries over to future measurements. One important self-similarity in psychological time series is \texit{inertia} --- how much a measurement carries over to its next measurement \citep{kuppens2010, suls1998}. If inertia is high a development tends to stay in a certain state. Because high inertia is resistant to change, in emotion dynamics high inertia of negative affect has been found to be indicative of under-reactive systems and to be characteristic of psychological maladjustment \citep{kuppens2010}. In a similar vein, high inertia in negative affect at baseline was even predictive of the initial onset of depression \citep{kuppens2012}. Conceptually, inertia is more broadly connected to research questions such as: ``Do patients stay in a negative mood for several measurements?''

@@

\paragraph{Linear trend.}
In non-stationary time series, a linear trend can be observed when there is a consistent increase or decrease in the data over time \citep{nyblom1986}. For psychological time series, researchers have, for example, pointed out the importance of linear trends in interpersonal communications \citep{vasileiadou2014}, and emotion dynamics \citep{oravecz2016}. Theoretically, linear trends are often considered the simplest way of assessing whether a psychological theory of change is appropriate \citep{gottman1969}. In empirical practice, linear trends are, thus, commonly exemplified by research questions such as "Do patient symptoms improve consistently?" or "Does worker productivity decline continuously?"

For the variables in our illustration data set, we chose an overall linear regression slope to capture the linear trend. The regression slope $b_{ij}$ provides the average change from one time point $t$ to the next across all measurements $x$ of a time series $X$ of participant $i$ and variable $j$. The specific form of the OLS slope formula we provide below calculates $b_{ij}$ as the sum across all time points of the product of the deviation of time $t$ from its mean $\overline{t}$ and the deviation of $x_{ij}$ from its mean $\overline{x}_{ij}$ at each time point, divided by the sum across all time points of the square of the deviation of time from its mean ($\sum(t-\overline{t})^2$). Intuitively, the formula captures the rate of change of variable $x_{ij}$ with respect to time. This slope will indicate how the variable $x_{ij}$ changes over time, controlling for its mean value and the mean of time. If the slope is positive, $x_{ij}$ increases over time; if it's negative, $x_{ij}$ decreases over time.

@@

\paragraph{Nonlinearity.}
Changes in psychology are not always linear, instead, nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. Conceptually, researchers might have research questions about the type of the development: "Is the development of anxiety a nonlinear process?" as well as the shape and structure of the development: "How many spikes in anxiety did a patient experience?" 
We summarized the nonlinear trend with the \textit{estimated degrees of freedom} of an empty GAM spline model. The $edf$ summarizes the \textit{wiggliness} of a spline trend line by estimating the number of parametric and non-parametric (or smooth) terms $p$ in the model. An edf of 1 would be equivalent to a linear relationship (i.e., one linear slope parameter), whereas a higher edf (particularly an edf > 2) is indicative of a non-linear trend.

@@

Beyond our main features of interest, we also extracted the participant's number of completed ESM measurements to ensure that the clusters are comparable in that regard (i.e., to exclude spurious explanations for the cluster assignments). After the feature extraction, we found that about `r format(round(pMissFull$pmiss %>% filter(set == "All") %>% select(pMissTotal) %>% pull, 2), nsmall=2)`% of the extracted features are missing across the `r ncol(featFull$features %>% select(-ends_with(featureExclusion)))-1` features per participant. This might, for example, happen if participants do not have two subsequent measurements with outgroup interactions, so that an autocorrelation with lag-1 cannot be calculated for the contact-specific variables. The small number of missing values indicates that the feature-based approach indeed largely avoids the structural missingness issue. However, even the few missing values can be an issue for some feature reduction or feature clustering algorithms. We, thus, impute the missing feature values with a single predictive mean matching imputation using the MICE library \citep[][]{buuren2011}. Note again that with this procedure we only need to impute an extremely small number of missing values as most feature calculations can use the available data instead.

### 2023-07-22 sections/04FeatureReduction.Rmd
feedback Rei

For our own illustration data we chose a feature projection method to reduce the dimensionality of our extracted features. We particularly chose the feature projection method for its broad applicability. We, specifically, selected the commonly used \textit{principal component analysis} (PCA). Some of the more tailor-made feature selection algorithms can be more accurate in reducing the feature dimensionality and might retain feature importance information more directly, depending on the specific data structure. However, PCAs have the distinct benefit that they are well-established within the psychometric literature \citep{jolliffe2011} and can broadly be applied to a wide variety of studies in an automatized manner \citep{abdi2010}. As our aim is to present a general illustration that can also be adopted across use-cases, we present the workflow using a PCA here but we encourage users to consider more specialized methods as well.

To use the principal component analysis with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA uses linear transformations in such a way that the first component captures the most possible variance of the original data \citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}. The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated \citep{shlens2014}. In practice, this meant that the PCA decomposed the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90% of the original variance \citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}. For our illustration we select the first `r pca_cutoff` principal components that explain 80% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`%). For the extracted principal components we save the `r pca_cutoff` PC-scores for each participant (i.e., the participants' coordinates in the reduced dimensional space).

We would like to comment on two practical matters when using principal components --- the amount of dimensionality reduction and the interpretation of the principal components. As for the expected dimensionality reduction, given its methodology, PCAs tend to 'work better' at reducing dimensions with (highly) correlated variables \citep[e.g.,][]{jolliffe2002}. Thus, with a set of very homogeneous variables and features users will need less principal components to explain a large amount of variance, while a more diverse set of variables and features will tend to require more principal components to capture the same amount of variance \citep[e.g.,][]{abdi2010}. Our `r pca_cutoff` principal components are still a relatively high number of variables but this is not surprising as we chose a diverse conceptualization and a diverse set of time series features. As for the interpretability, PCA allows users to extract information on the meaning of the principal components. In particular, because the principal components are linear combination of original features, users can extract the relative importance of each feature for the extracted principal components (i.e., the eigenvectors). While this can be useful in understanding the variance in the original data or help with manual feature selection, we use the PCA purely to reduce the dimensionality for the clustering step. Instead of relying on the principal components, we use the original features of interest to interpret the later extracted clusters. We particularly advocate for such an approach if all original features are considered meaningful in understanding the time series and users would like to retain the features for inpterpretation (irrespective of the features importance).

>>>>

For our own illustration data, we chose a feature projection method to reduce the dimensionality of our extracted features. We particularly chose the feature projection method for its broad applicability. We, specifically, selected the commonly used \textit{principal component analysis} (PCA). Some of the more tailor-made feature selection algorithms can be more accurate in reducing the feature dimensionality and might retain feature importance information more directly, depending on the specific data structure. However, PCAs have the distinct benefit that they are well-established within the psychometric literature \citep{jolliffe2011} and can broadly be applied to a wide variety of studies in an automatized manner \citep{abdi2010}. As our aim is to present a general illustration that can also be adopted across use cases, we present the workflow using a PCA here but we encourage users to consider more specialized methods as well.

To use the principal component analysis with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA uses linear transformations in such a way that the first component captures the most possible variance of the original data \citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}. The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated \citep{shlens2014}. In practice, this meant that the PCA decomposed the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90% of the original variance \citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}. For our illustration, we select the first `r pca_cutoff` principal components that explain 80% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`%). For the extracted principal components we save the `r pca_cutoff` principal component scores for each participant (i.e., the participants' coordinates in the reduced dimensional space; PC-scores).

We would like to comment on two practical matters when using principal components --- the amount of dimensionality reduction and the interpretation of the principal components. As for the expected dimensionality reduction, given its methodology, PCAs tend to 'work better' at reducing dimensions with (highly) correlated variables \citep[e.g.,][]{jolliffe2002}. Thus, with a set of very homogeneous variables and features users will need fewer principal components to explain a large amount of variance, while a more diverse set of variables and features will tend to require more principal components to capture the same amount of variance \citep[e.g.,][]{abdi2010}. Our `r pca_cutoff` principal components are still a relatively high number of variables but this is not surprising as we chose a diverse conceptualization and a diverse set of time series features. As for interpretability, PCA allows users to extract information on the meaning of the principal components. In particular, because the principal components are linear combinations of the original features, users can extract the relative importance of each feature for the extracted principal components (i.e., the eigenvectors). While this can be useful in understanding the variance in the original data or help with manual feature selection, we use the PCA purely to reduce the dimensionality for the clustering step. Instead of relying on the principal components, we use the original features of interest to interpret the later extracted clusters. We particularly advocate for such an approach if all original features are considered meaningful in understanding the time series and users would like to retain the features for interpretation (irrespective of the features' importance).

### 2023-07-22 sections/05FeatureClustering.Rmd
:memo: feat red adj
typos
feedback Rei

In our case we entered the participants' PC-scores from the feature reduction step into the k-means algorithm. Because we did know the underlying number of clusters within our sample, we calculated the cluster solutions for $k =$ 2 through 10. To avoid local minima we used 100 random initial centroid positions for each run. Each of the 9 cluster solutions converged within the iteration limit. In the next step we will then evaluate which of the extracted cluster solutions offers the best fit with the data.

>>>>

In our case, we entered the participants' PC-scores from the feature reduction step into the k-means algorithm. Because we did not know the underlying number of clusters within our sample, we calculated the cluster solutions for $k=\{2, \dots , 10\}$ . To avoid local minima we used 100 random initial centroid positions for each run. Each of the 9 cluster solutions converged within the iteration limit. In the next step, we will then evaluate which of the extracted cluster solutions offers the best fit with the data.
